Estimating duration in the presence of misclassification

### 2024-03-29 introduction.tex
merge new intro into old

COVID-19, the disease caused by the virus SARS-CoV-2, is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
Beyond the lives lost, there were massive health, economic, and societal costs.

History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current mean impact of pandemics is over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it uses the behaviour and immunity status of the population at time $t$.
With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
$\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the probability of transmission upon contact, and the proportion of the population that is susceptible.
Interventions act by reducing one of these components \eg masks reduce the probability of transmission and vaccination reduce the susceptible population.
Therefore, $\Re$ informs the strength of intervention required to control the epidemic.

The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections that would be generated by a single infection at time $t$ if in a fully susceptible population.
$\Rc$ is also known as the \emph{reproduction number excluding immunity}.
It describes the effect on the epidemic of interventions (except vaccination) that are in place at time $t$.
$\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: there are more individuals with immunity in January 2021, and the lockdowns had different measures in place.
$\Rc$ quantifies the latter effect.

These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
For example, effects due to seasonality, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.

Reproduction numbers are strongly connected to \emph{growth rates}.
If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially.
The rate of this exponential growth is its \emph{growth rate}, normally denoted $r$.
For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantages of using the growth rates are that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
Yet the size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
Incidence is translated into disease burden using a measure of severity; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
Estimation of severity is beyond the scope of this thesis.

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to inferring the parameters of this process.
The components of transmission (\eg the number of contacts each individual makes per day) allow interventions to be targeted.
They are also important for forecasting future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time.
Whether a given individual is \emph{prevalent} (has the disease) is often measurable through diagnostic testing.
Performing diagnostic tests in a sample of the population at a given point in time is known as a \emph{prevalence survey}.
Prevalence surveys can estimate the population prevalence but are typically very expensive to power appropriately.
Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).

@@

Collecting data that directly inform inference of the quantities in this section is expensive or impossible.
Hence, they are typically inferred from other data sources through modelling.

@@

Data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
These biases need to be accounted for when analysing the data.

Here I discuss three types of data which are useful for surveillance: incidence-type, prevalence-type, and data from cohort studies.
\emph{Incidence-type} and \emph{prevalence-type} data can inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
For example, the time between infection and symptom onset.
These quantities can be estimated from other studies, \eg cohort studies.

@@

Incidence-type data originate from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, particularly that the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

An example of incidence-type data are community testing, reliant on individuals presenting for a test.
However, individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.
Unfortunately, data to estimate these changes is normally unavailable.

Data on more severe symptoms are typically less affected by these biases.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}).
This means that the observations do not provide much information on incidence in those subgroups.
Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated with much precision~\autocite{swallow2022challenges}.

@@

Prevalence-type data originate from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
Commonly used disease states are the presence or value of a biomarker or symptoms.
For example, the number of hospital beds occupied by individuals with the disease or the proportion of the population with a marker of prior infection.
The issues associated with incidence-type data often also apply to prevalence-type data.

Prevalence surveys are one form of prevalence-type data that can be unbiased.
Here, a representative proportion of the population (or a subpopulation of interest) are tested for a marker of current infection.
For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.

However, \emph{seroprevalence} surveys are more feasible.
Seroprevalence is the proportion of individuals in the population which have antibodies to the disease, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.

\subsection{Cohort studies}

Other epidemiological and/or biological quantities are normally required to estimate transmission.
These depend on the model type, often including the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
Cohort studies (following a group of individuals over time), are a study design which can estimate many of these quantities.

@@

The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which incidence arises.
A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process~\autocite{beckerCOVIDmodels}.
Stronger assumptions allow more insight into the transmission process and lower the variance of the estimates but can introduce bias and be computationally challenging to fit.

The models with the weakest assumptions are \emph{phenomenological} models.
These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of inference.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
A common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
The renewal equation for an infectious disease~\autocite{fraserRenewal}, in the most general form, is:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
This partially describes the transmission process, hence the name semi-mechanistic.
The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Extensions of \cref{intro:eq:renewal} that track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex models, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, in particular to incorporate heterogeneity.
The age structure of the population is important for respiratory diseases because individuals meet individuals of their own age more frequently.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

@@

\emph{Disease models} link the transmission model to a latent quantity, $W_t$.
$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see next section).
For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.

At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellBayesian}.
This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that ever appear in $W_t$; and $f(i)$ is the probability that an infection at time $t-i$ is included in count $W_t$, conditional on it ever being in one of the counts.
$f$ is known as a \emph{delay function}.
$f$ is commonly the pdf of a distribution, or a transformation of it.

@@

Applying \cref{intro:eq:disease-model} requires the delay function, $f$.
Typically, this is either the time until an event or the duration of a disease stage.
Time series data on $W_t$ cannot normally identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.

Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
Modelling a biomarker uses more information per observation but requires stronger assumptions.

The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
An individual is in the disease state while the biomarker is above or below some threshold.
This approach often requires parametric assumption(s) such as a functional form of the biomarker's trajectory.
In addition, if individual-level variation is included, which is generally important, a distributional form of random effects is required.
This requires both sufficient observations of the biomarker per individual and an assumption on the form of the distribution.
More details are in \cref{E-ATACCC}.

Modelling the duration directly considers only the binary result at each test (positive or negative), for instance in a \emph{survival analysis} framework.
This approach generally requires fewer parameters reducing the computational burden.
It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
More details are in \cref{E-perf-test}.

Both cases require longitudinal data, suggesting that cohort studies are the most appropriate study design.

@@

The surveillance systems used in response to SARS-CoV-2 are historically unique.
Modern technology allowed large amounts of data to be collected, distributed, and analysed.
While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, it was far slower moving.
Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.

A deep understanding of each data source's potential was not possible in real-time.
In this thesis, I will investigate the potential of the prevalence surveys in greater detail.
The disease model will be a prevalence model (defined further in \cref{E-inc-prev:sec:prevalence-process}) with the delay function being the duration of PCR positivity.

@@

The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing (known as pillar 2 testing), the number of severe events, seroprevalence surveys, and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.

@@

For large parts of the pandemic, community testing was widely available in the UK.
However, it did not test a representative sample of the population.
The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
These include: the policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.

Severe events counts are generally a more reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context.
Therefore, these data are subject to fewer biases than community testing.
However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only affecting younger individuals.

@@

Seroprevalence surveys were conducted throughout the epidemic, the largest of which was tested donated blood~\autocite{amirthalingamSeroprevalence}.
In addition to the normal problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and observe the whole population.
Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
Despite this, they are considered the gold-standard metric.

The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal study, testing a large number of individuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

A major advantage of the CIS over REACT is that the CIS had continual data collection, while REACT has no test results between rounds.
The lack of data between rounds meant that using REACT data to infer transmission in these periods requires strong model assumptions.
A particular instance of this is that a peak in late December 2020 occurred between rounds of the REACT survey, complicating analysis of the survey data at that time~\autocite{rileyREACTround8}.

\subsubsection{Cohort studies}

>>>>

Infection with the SARS-CoV-2 virus is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
Beyond the lives lost, there were massive health, economic, and societal costs.

History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
\begin{figure}
    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
    \caption[Timeline of pandemics.]{%
        Timeline of pandemics.
        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
    }
    \label{intro:fig:pandemic-timeline}
\end{figure}
Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
Recent estimates suggest that the current impact of pandemics is, on average, over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.

@@

The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
In particular, it includes the effects of the behaviour and immunity status of the population at time $t$.
With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
$\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the probability of transmission upon contact, and the proportion of the population that is susceptible.
Interventions act by reducing one of these components \eg masks reduce the probability of transmission and vaccination reduce the susceptible population.
Therefore, $\Re$ informs the strength of intervention required to control the epidemic.

The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections that would be generated by a single infection at time $t$ if in a fully susceptible population.
$\Rc$ is also known as the \emph{reproduction number excluding immunity}.
It describes the effect on the epidemic of interventions (except vaccination) that are in place at time $t$.
$\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.

% To see the differing uses of the reproduction numbers, consider the difference in the epidemic in England in March 2020, just following the start of the first lockdown, and in January 2021 just following the start of the third lockdown.
% These times have (at least) three epidemiologically relevant differences: (1) in January 2021 the Alpha variant was circulating which is more transmissible than the original variant present in March 2020; (2) the population behaved differently because the exact meaning of ``lockdown'' differed and attitudes had changed; (3) in January 2021, some individuals had immunity due to either prior infection or the start of the vaccination roll-out.
% The original and Alpha variants have different values for $\R$, isolating effect 1.
% $\Rc$ differs during the two periods, including effects 1 and 2.
% $\Re$ includes all these effects, as well as any others.

% These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
% For example, how to average across seasonal effects and a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
% However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.

Reproduction numbers are strongly connected to \emph{growth rates}.
If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially, with a growth rate $r$.
For a given disease and population, a growth rate can be defined corresponding to each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
The advantages of using the growth rates are that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.

@@

Growth rates and reproduction numbers only provide information on how an epidemic is changing.
The size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.

\emph{Incidence} is the number of new infections in a given time period (see \cref{E-inc-prev:sec:definitions} for a formal definition).
It is the primary metric for the magnitude of a pandemic .
Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths); although this requires a severity measure, such as the IFR (infection fatality rate), the proportion of infected individuals who die.

\emph{Transmission} refers to the process by which an infectious individual generates further infections.
Estimating transmission refers to inferring this process.
Understanding this process, \eg the number of contacts each individual makes per day, allows targeting interventions and forecasting future incidence.

\emph{Prevalence} is the proportion of the population with the disease at a given time (see \cref{E-inc-prev:sec:definitions} for a formal definition).
Whether a given individual is \emph{prevalent}, \ie has the disease, is often measurable through diagnostic testing.
Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).

@@

Collecting data that \emph{directly} inform inference of the quantities in this section is challenging or prohibitively expensive.
Hence, they are typically estimated from other data sources through appropriate statistical methods~\autocite{heldHandbook,oneillIntroduction}.

@@

Data sources to inform outbreak surveillance are often observational in nature, \ie they are not the output of designed studies.
Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
These biases need to be accounted for when analysing the data.

In what follows, I discuss three types of data: incidence-type, prevalence-type, and data from longitudinal studies.
\emph{Incidence-type} and \emph{prevalence-type} data can inform estimation of incidence and transmission.
However, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities, typically estimated from longitudinal studies.
For example, the time between infection and symptom onset.

@@

Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
This type of data typically suffers from reporting biases, most problematically that the proportion of infections detected varying over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.

An example of incidence-type data are the number of cases identified through community testing, which relies on individuals presenting for a test.
Individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.

Data on severe disease is typically less affected by biases.
An extreme example is the development of AIDS following infection by HIV.
Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.

However, severe disease can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}), meaning they rarely appear in this data.
Therefore, the data does not provide much information on incidence in those subgroups.
Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated with much precision~\autocite{swallow2022challenges}.

@@

Prevalence-type data originate from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
For example, the number of individuals in hospital with the disease or the proportion of the population with a marker of prior infection.
The same issues associated with incidence-type data often also apply to prevalence-type data.

Prevalence can be estimated through a \emph{prevalence survey}.
A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately.
On the other hand, they are a source of prevalence-type data that can be unbiased.
For instance, testing a representative proportion of the population (or a subpopulation of interest) for a marker of current infection.

However, \emph{seroprevalence} surveys are more feasible.
Seroprevalence is the proportion of individuals in the population which have antibodies against a particular infection, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.

\subsection{Longitudinal studies}

Other epidemiological and/or biological quantities are normally required to estimate transmission.
These include the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
Longitudinal studies, that follow the same individuals over time, allow estimation of many of these quantities.

@@

The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which infections are generated.
A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process~\autocite{beckerCOVIDmodels}.

The models with the weakest assumptions are \emph{phenomenological} models.
These make almost no assumptions about the transmission process, viewing incidence as resulting from a counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
A common model is the Poisson process, and the intensity of the process is the target of inference.
% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.

Next, are semi-mechanistic models.
These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal} of the most form:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
    \label{intro:eq:renewal}
\end{align}
where $Z(t)$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
The model parameters can be used to define an effective reproduction number as either $\Re = \int_0^\infty \beta(t, \tau) d\tau$ or $\Re = \int_0^\infty \beta(t+\tau, \tau) d\tau$~\autocite{gosticPractical}.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{thompsonImproved,gosticPractical}.

Finally, mechanistic models are the most complex models, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic (how it changes over time) are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, incorporating more flexible definitions and/or heterogeneity (see \cref{SEIR:sec:mechanistic-models}).
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

@@

The infection process is hardly ever directly observed; however, we often observe a second process.
\emph{Disease models} link the transmission model to the value, $W_t$ of this second process at time $t$.
For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.
$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see next section), which could introduce additional complications.

At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellBayesian}.
This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in $W_t$; and, conditional on being observed, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
$f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf of a distribution, or a transformation of it.

@@

Applying \cref{intro:eq:disease-model} requires knowledge of the delay function, $f$.
Typically, this is either the time to an event or the duration of a disease stage.
Time series data on $W_t$ cannot identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}, most commonly longitudinal studies.

Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly, for instance, in a \emph{survival analysis} framework~\autocite[e.g.][]{sweetingEstimating}.

The biomarker approach models the trajectory of a biomarker over time.
Applying a threshold to the biomarker defines a disease state.
This approach often requires parametric assumption(s) such as a functional form of the biomarker's trajectory.
In addition, including individual-level variation, which is generally important, requires sufficient observations of the biomarker per individual and an assumption on the form of the distribution of the variation.
More details are in \cref{E-ATACCC}.

Modelling the duration directly considers only the binary result at each test (positive or negative).
This approach generally requires fewer parameters reducing the computational burden.
It is also easier to specify non-parametric forms of the distribution, which is useful in situation where the tail behaviour may be important.
% A more detailed review of the literature is in the relevant chapters.
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
More details are in \cref{E-perf-test}.

@@

The response to the SARS-CoV-2 pandemic in the UK has been unprecedented, both in terms of the scientific input to policy and the surveillance data made available.
The most prominent surveillance outputs were produced by members of SPI-M-O (the Scientific Pandemic Influenza Group on Modelling Operational sub-group), a sub-committee of SAGE (the Scientific Advisory Group for Emergencies)~\autocite{medleySPIM,govSPIMO}.
SPI-M-O produced weekly estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
The committee peer-reviewed these estimates to provide quality assurance and formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
The consensus value was published on the Government's website and informed its response to the pandemic~\autocite{govRnumber}.

The nowcasting estimates were based on mechanistic and semi-mechanistic models with a wide variety of data sources.
The mechanistic models, in  particular, commonly incorporated multiple data sources.
The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).

Additionally, these models were used to explore policy scenarios.
For example, the decision at each of stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
These aided decision-makers in understanding plausible impacts of their actions.

@@

The surveillance systems used in response to SARS-CoV-2 are unique. 
The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing; severe events; seroprevalence and prevalence surveys.
Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.

@@

For large parts of the pandemic, community testing was widely available in the UK.
However, the number of tests carried out varied over time and space, and for a variety of reasons.
Reasons included: policy changes on who was eligible for testing; availability of tests; and willingness of individuals to be tested.
Therefore, changes in the number of positive tests did not necessarily reflect changes in disease incidence.
The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although the lack of interpretability remained.

Severe events counts were generally a more reliable measure of the burden of the disease.
The criteria for being a severe event are normally well-defined and generally well-recorded.
Therefore, these data were subject to fewer biases than the cases detected through community testing.
However, SARS-CoV-2 rarely led to severe events in younger individuals, so severe event counts were insensitive to changes in transmission only affecting younger individuals.

@@

Seroprevalence surveys were conducted throughout the epidemic, the largest of which tested donated blood~\autocite{amirthalingamSeroprevalence}.
In addition to the problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.

The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
These were designed to avoid the biases of community testing and attempt to observe the general population.
These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
The CIS was a longitudinal household study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).

A second prevalence survey, known as REACT (the REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}, consisted in a series of cross-sectional surveys.
REACT conducted 19 surveys between May 2020 and March 2022, each including at least \numprint{100000} randomly sampled individuals from the UK, who contributed swabs for RT-PCR testing and questionnaire data~\autocite{elliottTwin}.

A major advantage of the CIS over REACT is that the CIS had continual data collection, while REACT has no test results between rounds.
This meant that using REACT data to infer transmission in these periods requires strong model assumptions.
A particular instance of this is that a peak in prevalence in late December 2020 occurred between rounds of the REACT survey, complcating any analysis of REACT data over that period~\autocite{rileyREACTround8}.

\subsubsection{Longitudinal studies}

Many longitudinal studies were conducted throughout the pandemic.
They aimed to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus.
These include the CIS, previously mentioned.
% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
